Intersection Types, Sub-formula Property, and the Functional Character of the Lambda Calculus

### 2018-08-09 DenotSem/determ.tex
check

I am particularly interested in this theorem because $\beta$-equality
can be obtained as a corollary.
\[
\ESEM{(\lam{x}e)\app e'}\rho = \ESEM{[x{:=}e']e}\rho
\]
This would enable the modeling of the call-by-name $\lambda$-calculus
and it would also enable the use of $\beta$-equality in a
call-by-value setting when $e'$ is terminating (instead of restricting
$e'$ to be a syntactic value).

@@

Turning back to the Join Theorem, I restated it in terms of the
intersection type system. Instead of using the letter $\rho$ for
environments, we shall switch to $\Gamma$ because they now contain
types instead of values. \\

\noindent \textbf{Theorem} (Intersection Introduction)\\
  If $\Gamma \vdash e : A$, $\Gamma' \vdash e : B$, and $\Gamma \sim
  \Gamma'$, 
  then $\Gamma\sqcap\Gamma' \vdash e : A \sqcap B$. \\

@@

We shall represent variables with de Bruijn indices, so an environment
$\Gamma$ is a sequence of types.  The denotational semantics of the CBV
$\lambda$-calculus is defined as follows.
\begin{align*}
  \ESEM{n}\Gamma &= \UP{n} \\
  \ESEM{x}\Gamma &=  \IF x < |\Gamma| \THEN \UP{\Gamma[k]} \ELSE \emptyset \\
  \ESEM{\lambda e }\Gamma &= \{ A \mid \WF{A} \text{ and } \FSEM{A}(e,\Gamma) \} \\
  \ESEM{e_1\app e_2}\Gamma &= \left\{ C\, \middle| \exists A,B.\;
      \begin{array}{l}
      A \in \ESEM{e_1}\Gamma,
      B \in \ESEM{e_2}\Gamma,\\
      A <: B \to C, \text{ and } \WF{C}
      \end{array}
        \right\} \\
  \ESEM{f(e_1,e_2)}\Gamma &=
      \left\{ C\, \middle| \begin{array}{l}
       \exists A,B,n_1,n_2.\; A \in \ESEM{e_1}\Gamma, B \in \ESEM{e_2}\Gamma,\\
      \qquad A <: n_1, B <: n_2, \SEM{f}(n_1,n_2) <: C, \WF{C} 
      \end{array} \right\} \\
  \ESEM{\IF e_1 \THEN e_2 \ELSE e_3}\Gamma &=
    \left\{ B\, \middle| \exists A.\;
    \begin{array}{l}A \in \ESEM{e_1}\Gamma, A <: n,\\
           n = 0 \Rightarrow B \in \ESEM{e_3}\Gamma,\\
           n \neq 0 \Rightarrow B \in \ESEM{e_2}\Gamma
    \end{array}
    \right\}
    \\[2ex]
    \intertext{where}
   \FSEM{n}e\Gamma &= \mathit{false} \\
   \FSEM{A \sqcap B}e \Gamma &= \FSEM{A}e\Gamma \text{ and } \FSEM{B}e\Gamma\\
   \FSEM{A \to B}e \Gamma &= B \in \ESEM{e} (A, \Gamma)
\end{align*}

@@

Now for the main event, the proof of the Join Theorem!  Though we'll
rebrand it as the Meet Theorem, given our switch from values to
intersection types.\\

\noindent \textbf{Theorem} (Meet)\\
If $A_1 \in \ESEM{e}\Gamma_1$, $A_2 \in \ESEM{e}\Gamma_2$,
both $\Gamma_1$ and $\Gamma_2$ are well formed,
and $\Gamma_1 \sim \Gamma_2$, \\
then $A_1 \sqcap A_2 \in \ESEM{e}(\Gamma_1\sqcap\Gamma_2)$. \\[1ex]
%
\textbf{Proof}
We proceed by induction on $e$.
\begin{itemize}
\item Case $e=k$ ($k$ is a de Bruijn index for a variable):\\
  We have $\Gamma_1[k] <: A_1$ and $\Gamma_2[k] <: A_2$,
  so $\Gamma_1[k] \sqcap \Gamma_2[k] <: A_1 \sqcap A_2$.
  Also, because $\Gamma_1 \sim \Gamma_2$ we have
  $\Gamma_1[k] \sim \Gamma_2[k]$ and therefore
  $A_1 \sim A_2$, by the Consistency and Subtyping Lemma.
  So we have $\WF{A_1 \sqcap A_2}$ and this case is finished.

@@

\item Case $e=\lambda e$: \\

@@

>>>>

\noindent I am particularly interested in this theorem because
$\beta$-equality can be obtained as a corollary.
\[
\ESEM{(\lam{x}e)\app e'}\rho = \ESEM{[x{:=}e']e}\rho
\]
This would enable the modeling of the call-by-name $\lambda$-calculus
and it would also enable the use of $\beta$-equality in a
call-by-value setting when $e'$ is terminating (instead of restricting
$e'$ to be a syntactic value).

@@

Turning back to the Join Theorem, I restated it in terms of the
intersection type system and rebranded it the Meet Theorem. Instead of
using the letter $\rho$ for environments, we shall switch to $\Gamma$
because they now contain types instead of values. \\

\noindent \textbf{Theorem} (Meet)\\
  If $\Gamma \vdash e : A$, $\Gamma' \vdash e : B$, and $\Gamma \sim
  \Gamma'$, 
  then $\Gamma\sqcap\Gamma' \vdash e : A \sqcap B$. \\

@@

We shall represent variables with de Bruijn indices, so an environment
$\Gamma$ is a sequence of types.  The denotational semantics of the CBV
$\lambda$-calculus is defined as follows.
\begin{align*}
  \ESEM{n}\Gamma &= \UP{n} \\
  \ESEM{x}\Gamma &=  \IF x < |\Gamma| \THEN \UP{\Gamma[k]} \ELSE \emptyset \\
  \ESEM{\lambda e }\Gamma &= \{ A \mid \WF{A} \text{ and } \FSEM{A}e\Gamma \} \\
  \ESEM{e_1\app e_2}\Gamma &= \left\{ C\, \middle| \exists A,B.\;
      \begin{array}{l}
      A \in \ESEM{e_1}\Gamma,
      B \in \ESEM{e_2}\Gamma,\\
      A <: B \to C, \text{ and } \WF{C}
      \end{array}
        \right\} \\
  \ESEM{f(e_1,e_2)}\Gamma &=
      \left\{ C\, \middle| \begin{array}{l}
       \exists A,B,n_1,n_2.\; A \in \ESEM{e_1}\Gamma, B \in \ESEM{e_2}\Gamma,\\
      \qquad A <: n_1, B <: n_2, \SEM{f}(n_1,n_2) <: C, \WF{C} 
      \end{array} \right\} \\
  \ESEM{\IF e_1 \THEN e_2 \ELSE e_3}\Gamma &=
    \left\{ B\, \middle| \exists A.\;
    \begin{array}{l}A \in \ESEM{e_1}\Gamma, A <: n,\\
           n = 0 \Rightarrow B \in \ESEM{e_3}\Gamma,\\
           n \neq 0 \Rightarrow B \in \ESEM{e_2}\Gamma
    \end{array}
    \right\}
    \\[2ex]
    \intertext{where}
   \FSEM{n}e\Gamma &= \mathit{false} \\
   \FSEM{A \sqcap B}e \Gamma &= \FSEM{A}e\Gamma \text{ and } \FSEM{B}e\Gamma\\
   \FSEM{A \to B}e \Gamma &= B \in \ESEM{e} (A, \Gamma)
\end{align*}

@@

Now for the main event, the proof of the Meet Theorem! \\

\noindent \textbf{Theorem} (Meet)\\
If $A_1 \in \ESEM{e}\Gamma_1$, $A_2 \in \ESEM{e}\Gamma_2$,
both $\Gamma_1$ and $\Gamma_2$ are well formed,
and $\Gamma_1 \sim \Gamma_2$, \\
then $A_1 \sqcap A_2 \in \ESEM{e}(\Gamma_1\sqcap\Gamma_2)$ and
  $\WF{A_1 \sqcap A_2}$. \\[1ex]
%
\textbf{Proof}
We proceed by induction on $e$.
\begin{itemize}
\item Case $e=k$ ($k$ is a de Bruijn index for a variable):\\
  We have $\Gamma_1[k] <: A_1$ and $\Gamma_2[k] <: A_2$,
  so $\Gamma_1[k] \sqcap \Gamma_2[k] <: A_1 \sqcap A_2$.
  Also, because $\Gamma_1 \sim \Gamma_2$ we have
  $\Gamma_1[k] \sim \Gamma_2[k]$ and therefore
  $A_1 \sim A_2$, by the Consistency and Subtyping Lemma.
  So we have $\WF{A_1 \sqcap A_2}$ and this case is finished.

@@

\item Case $e=\lambda e$: \\
  We need to show that $\WF{A_1 \sqcap A_2}$ and
  $\FSEM{A_1 \sqcap A_2}e(\Gamma_1\sqcap\Gamma_2)$.
  For the later, it suffices to show that $A_1 \sim A_2$,
  which we shall do by showing that their atoms are consistent.
  Suppose $A'_1 \in \ATOMS{A_1}$ and $A'_2 \in \ATOMS{A_2}$.
  Because $\FSEM{A_1}e\Gamma_1$ 
  we have $A'_1 =A'_{11} \to A'_{12}$ and
  $A'_{12} \in \ESEM{e}(A'_{11},\Gamma_1)$. Similarly,
  from $\FSEM{A_2}e\Gamma_2$ we have
  $A'_2 =A'_{21} \to A'_{22}$ and $A'_{22} \in \ESEM{e}(A'_{21},\Gamma_2)$.
  We proceed by cases on whether $A'_{11} \sim A'_{21}$.
  \begin{itemize}
  \item Sub-case $A'_{11} \sim A'_{21}$:\\
    By the induction hypothesis, we have
    $\WF{A'_{12} \sqcap A'_{22}}$ from which we
    have $A'_{12} \sim A'_{22}$ and therefore
    $A'_{11}\to A'_{12} \sim A'_{21} \to A'_{22}$.

    %% \[
    %%   A'_{12} \sqcap A'_{22} \in
    %%   \ESEM{e}(A'_{11}\sqcap A'_{21},\Gamma_1 \sqcap \Gamma_2)
    %%   \quad\text{and} \quad
    %%   \WF{A'_{12} \sqcap A'_{22}}
    %% \]

  \item Sub-case $A'_{11} \not\sim A'_{21}$:\\
      It immediately follows that $A'_{11}\to A'_{12} \sim A'_{21} \to A'_{22}$.
  \end{itemize}

  It remains to show $\FSEM{A_1 \sqcap A_2}e(\Gamma_1\sqcap\Gamma_2)$.
  This follows from two uses of the Weakening Lemma to obtain
  $\FSEM{A_1}e(\Gamma_1\sqcap\Gamma_2)$ and
  $\FSEM{A_2}e(\Gamma_1\sqcap\Gamma_2)$.

\item Case $e = (e_1 \app e_2)$: \\
  We have
  \[
  B_1 \in \ESEM{e_1}\Gamma_1 \quad
  C_1 \in \ESEM{e_2}\Gamma_1 \quad
  B_1 <: C_1 \to A_1 \quad
  \WF{A_1}
  \]
  and
  \[
  B_2 \in \ESEM{e_1}\Gamma_2 \quad
  C_2 \in \ESEM{e_2}\Gamma_2 \quad
  B_2 <: C_2 \to A_2 \quad
  \WF{A_2}
  \]
  By the induction hypothesis, we have
  \[
  B_1 \sqcap B_2 \in \ESEM{e_1}(\Gamma_1 \sqcap \Gamma_2)
  \quad
  \WF{B_1 \sqcap B_2}
  \]
  and
  \[
  C_1 \sqcap C_2 \in \ESEM{e_2}(\Gamma_1 \sqcap \Gamma_2)
  \quad
  \WF{C_1 \sqcap C_2}
  \]
  We obtain $A_1 \sim A_2$ by the Application Consistency Lemma,
  and then by the Application Intersection Lemma we have
  \[
    B_1 \sqcap B_2 <: (C_1 \sqcap C_2) \to (A_1 \sqcap A_2)
  \]
  So we have $A_1 \sqcap A_2 \in \ESEM{e}(\Gamma_1 \sqcap \Gamma_2)$.

  Also, from $A_1 \sim A_2$, $\WF{A_1}$, and $\WF{A_2}$, we conclude
  that $\WF{A_1 \sqcap A_2}$.
\item Case $e= f(e_1,e_2)$: \\
  (This case is not very interesting. See the Isabelle proof
   for the details.)

 \item Case $e= \IF e_1 \THEN e_2 \ELSE e_3$: \\
  (This case is not very interesting. See the Isabelle proof
   for the details.)

@@

I thought that the following Subsumption Theorem would be needed to
prove the Meet Theorem, but it turned out not to be necessary, which
is especially nice because the proof of the Subsumption Theorem turned
out to depend on the Meet Theorem!\\

\noindent \textbf{Theorem} (Subsumption)\\
If $A \in \ESEM{e}\Gamma$, $A <: B$, and both $B$ and $\Gamma$ are
well-formed, then $B \in \ESEM{e}\Gamma$. \\

\noindent The proof is by induction on $e$ and all but the case
$e=\lambda e'$ are straightforward. For that case, we use the
following lemmas.\\

\noindent \textbf{Lemma} (Distributivity for $\mathcal{F}$)\\
If $\FSEM{(A \to B)\sqcap (C \to D)} e \Gamma$,
$A \sim C$, and everything is well formed, then
$\FSEM{(A\sqcap C) \to (B\sqcap D)} e \Gamma$.\\
(The proof is direct, using the Meet Theorem
and the Weakening Lemma.)\\

\noindent \textbf{Lemma} ($\mathcal{F}$ and Intersections)\\
Suppose $\Gamma_1$ is a non-empty sequence of well-formed and
consistent function types.
If $\FSEM{\bigsqcap \Gamma_1} e \Gamma_2$, then
$\FSEM{\bigsqcap\mathrm{dom}(\Gamma_1) \to \bigsqcap \mathrm{cod}(\Gamma_1)}
e \Gamma_2$. \\
(The proof is by induction on $\Gamma_1$ and uses the previous lemma.)

\section{Conclusion}

This result can be viewed a couple ways. As discussed at the beginning
of this post, establishing the Meet Theorem means that the this
call-by-value denotational semantics respects $\beta$-equality for any
terminating argument expression. This is useful in proving the
correctness of a function inlining optimizer. Also, it would be
straightforward to define a call-by-name (or need) version of the
semantics that respects $\beta$-equality unconditionally.

Secondly, from the viewpoint of intersection type systems, this result
shows that, once we require types to be well formed (i.e. self
consistent), we no longer need the intersection introduction rule
because it is a consequence of having the subtyping rule for
distributing intersections through function types.
